---
title: "Bank Marketing Case Study"
author: "Austin Vanderlyn ajl745"
date: "2/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Executive Summary

Blah Blah Blah


### 2. The Problem

We have been given a dataset by the State College of Pennsylvania which collected 453 samples from people traveling from a location in the suburbs into the city. There are three different routes into the city, and the problem we need to solve is to predict the likelihood that a commuter will pick a specific route, a 4 lane arterial route. In addition, we would like to further identify which variables seem to most affect the commuter's choice of route. 

To accomplish this, 


### Appendix

Load Libraries
```{r}
install.packages("ROCR")
install.packages("lmtest")
library(logistf)
library(tidyverse)
library(car)
library(broom)
library(DescTools)
library(ROCR)
library(lmtest)
```

Load data
```{r}
BankData = read.csv("C:/Users/austi/OneDrive/Desktop/UTSA/UTSA Spring 2022/Applications/Data/bank-additional.csv")
str(BankData)
```

Character variables need to be converted to factor
```{r}
BankData$job = as.factor(BankData$job)
BankData$marital = as.factor(BankData$marital)
BankData$education = as.factor(BankData$education)
BankData$default = as.factor(BankData$default)
BankData$housing = as.factor(BankData$housing)
BankData$loan = as.factor(BankData$loan)
BankData$contact = as.factor(BankData$contact)
BankData$month = as.factor(BankData$month)
BankData$day_of_week = as.factor(BankData$day_of_week)
BankData$poutcome = as.factor(BankData$poutcome)
BankData$y = as.factor(BankData$y)
str(BankData)
```


Fit initial model;
```{r}
m1 = glm(y ~ ., data = BankData)
```



